---
title: 'Week 4: Overview of Local Networks'
author: "Ryan Light with significant collaboration with jimi adams"
date: "April 21, 2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Node Degree

Node degree is the number of edges adjacent to a node. It is a basic measure of connectivity and "importance." In directed graphs may consider *in-degree* the number of edges sent TO a node and *out-degree* the number of edges sent FROM a node or the total degree which is their sum.

We can find degree in igraph and we can also turn to our adjacency matrix to see how this works.

degree(g) creates a nodal undirected (or total) degree vector.

```{r results='hide', echo=TRUE, message=FALSE}
library(igraph)
valente <- read.graph("valente.net", format="pajek")
gender <- as.matrix(read.table("valente.clu", skip=1))	
V(valente)$gender <- as.vector(gender)					
is.directed(valente)
v_deg <- degree(valente)								
```
## Node Degree: In- and Out-degree

In-degree:
degree(g, mode="in")

Out-degree:
degree(g, mode="out")

```{r message=FALSE, echo=TRUE}
v_indeg <- degree(valente, mode="in")					
v_outdeg <- degree(valente, mode="out")				
#Sum to check
v_deg==v_indeg + v_outdeg								
```

## Degree Distribution

Let's see how degree is distributed in the Valente data by plotting the distribution using ggplot2. Ggplot2 is helpful for a bunch of graphic needs notes that there is a recipe/grammar to constucting graphs. 

You begin with data and add the "aesthetic mapping" or what you want to plot and then add layers and other specifications.

A generic histogram might look like this:

ggplot(data=df, aes=(variable)) + geom_histogram()


```{r message=FALSE, echo=TRUE}
library(ggplot2)

valente.degree <- data.frame(v_deg, v_indeg, v_outdeg)

ggplot(data=valente.degree, aes(v_deg)) + 
  
  geom_histogram(breaks=seq(2,14, by=1),
                col="black", 
                fill="black", 
                alpha = .2)+
 labs(title="Total Degree Distribution") +
  labs(x="Degree", y="Count")
```

We can look at the distribution of in-degree.

```{r }
ggplot(data=valente.degree, aes(v_indeg))  +
  geom_histogram(breaks=seq(-1,14, by=1), col="black", fill="black", alpha = .2)+
 labs(title="In-Degree Distribution") +
  labs(x="In-Degree", y="Count")
```

And the distribution of out-degree.

```{r message=FALSE, echo=FALSE}

ggplot(data=valente.degree, aes(v_outdeg)) + geom_histogram(breaks=seq(-1,14, by=1),
                                                          col="black", 
                                                          fill="black", 
                                                          alpha = .2)+
 labs(title="Out-Degree Distribution") +
  labs(x="Out-Degree", y="Count")
```

We can compare in-degree and out-degree.

```{r message=FALSE, echo=TRUE, fig.show='hide'}
ggplot(data=valente.degree,aes(v_indeg,v_outdeg))+
  geom_point(aes(colour=factor(V(valente)$gender+1)))+
  geom_smooth(method='lm', level=.95, formula=y~x)+
  theme(panel.background = element_blank())
```

##Network Composition

We can construct a variety of measures of network composition. For example we might just want the percentage of alters with a particular characteristics.

This is simply: (the # of alters with x)/degree. The mean of this simple composition is often reported for the entire network/data set. You can use matrices to construct these measures. 

For gender and the Valente data it would look like this: ADJMAT[32,32] X GENDER[32,2] = RECEIVE&SENT[32,2] 

##Network Composition

```{r , echo=TRUE, results="hide"}

girl <- ifelse(gender==1,1,0)
boy <- ifelse(gender==0, 1, 0)

gendermat <- cbind(girl, boy)

multiplication <- adjmat %*% gendermat

percent.cat <- data.frame(toboy=multiplication[,2], togirl=multiplication[,1], gender=gender, outgree=v_outdeg)

percent.cat$prop.boy <- percent.cat$toboy/percent.cat$outgree

percent.cat$prop.girl <- percent.cat$togirl/percent.cat$outgree

mean(percent.cat$prop.boy, na.rm=TRUE)

mean(percent.cat$prop.girl, na.rm=TRUE)
```
##Network Composition: Mixing Matrix

To see how groups send and receivev to each other you multiple the transpose of the attribute matrix by the result of the equation generated in the previous slide or,

*T(GENDER[32,2]) x RECEIVE&SENT[32,2] = MIXING MATRIX[2,2]*

```{r, echo=TRUE}
gendermixing <- t(gendermat) %*% multiplication
print(gendermixing)
```

##Network Composition: IQV

Index of Qualitative Variation measure variation among nominal variables. Here, it captures the diversity across ego's alters on a given nominal characteristic.  

Following Marsden (1985) and many others we use Argesti and Argesti (1978:208):

*(1-Sum(Category Proportions)^2)(# of Categories/1-# of Categories)*

```{r, echo=TRUE}
source("stanford_iqv.r") 
#stanford function...see https://sna.stanford.edu/lab.php?l=2
v_iqv_gender <- as.data.frame(get_iqvs(valente, "gender"))
quantile(v_iqv_gender, na.rm=TRUE)

```


##Network Density

Network density provides an indication of the strong ties within the network.

It is the number of ties/number of possible ties.

```{r, echo=TRUE}

graph.density(valente)

valente_ego_dens <- make_ego_graph(valente, 1) %>%
vapply(graph.density, numeric(1))

head(valente_ego_dens)

```

##Transitivity

Transitivity is a measure of local clustering. It consists of the proportion of closed triads given all possible triads. The igraph default is for each node but you can also calculate for the whole graph. This is an indication of how much local clustering is occuring at the graph as a whole.

```{r, echo=TRUE}

transitivity(valente)

transitive.valente <- transitivity(valente, type="local")
ego.size <- ego_size(valente)

tr.df <- data.frame(transitivity=transitive.valente, size=ego.size)

```
##Transitivity

```{r message=FALSE, echo=FALSE}
ggplot(data=tr.df,aes(transitivity,size))+
  geom_point()+
  geom_smooth(method='lm', level=.95, formula=y~x)+
  theme(panel.background = element_blank())
```

##Transitivity

```{r message=FALSE, echo=FALSE}
plot.igraph(valente, vertex.size=transitive.valente*20, edge.arrow.size=.5)
```

##Constraint

Another local measure is the extent to which an actor may be constrained by local clustering. Constraint captures the extent to which a node's local network is redundant. In other words, small networks with high redundancy will have high constraint scores.

"Network constraint measures the extent to which your time and energy are concentrated in a single group" (Burt)

```{r, echo=TRUE}

constraint.valente <- constraint(valente)
ego.size <- ego_size(valente)

con.df <- data.frame(constraint=constraint.valente, size=ego.size)

```

##Constraint

```{r message=FALSE, echo=FALSE}
ggplot(data=con.df,aes(constraint,size))+
  geom_point()+
  geom_smooth(method='lm', level=.95, formula=y~x)+
  theme(panel.background = element_blank())
```

##Constraint

```{r message=FALSE, echo=FALSE}
plot.igraph(valente, vertex.size=constraint.valente*20, edge.arrow.size=.5)
```
