---
title: 'Week 4: Overview of Local Networks'
output:
  html_document:
    df_print: paged
---

While many social scientists are easily seduced by the massive networks that we can observe and analyze, local networks remain a central concern of social scientists. As Smith (2020) describes, there are at least five reasons why local or ego networks remain important: 

1. Ego network data is easier to collect than whole network data.
2. Ego network data can be included in a "traditional" survey.
3. Ego network data pair nicely with other network data collection strategies and can inform those strategies.
4. Ego network data collection is amenable to sensitive topics, like sexual relationships or drug use.
5. Ego netework data is often detailed including a lot of information on both egos and their alters.

There are several ways that we can bring ego network data into R for analysis. Much of the data may look like a standard data frame: respondent/ego ID, alter1ID...Alter5ID, Alter1_Race...Alter5...Race, Alter1-Alter2_Edge...Alter4-Alter5_Edge.
Many statistics can be developed by summing over the alter characteristics and dividing by network size or by indexing the edge columns.

SNA scholars have also developed packages to simplify these tasks although the data structures are not always obvious. egor is a recent package that looks particularly promising for these tasks and is derived from the deprecated egonetR that was popular (Note the main importing functions are onefile_to_egor, twofile_to_egor, threefile_to_egor that each have different data structure expectations.)

For this tutorial we use igraph to observe local network characteristics within a whole graph. We also load the netseg pacakge for mixing matrices. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r results='hide', echo=TRUE, message=FALSE}
library(igraph)
library(netseg)
```

## Node Degree
Node degree is the most basic and one of the most widely used network measures. It is inherently a local network measure and each node has a degree score, but we can also talk about the average degree for the network as a whole.

Node degree is the number of edges adjacent to a node. It is a basic measure of connectivity and "importance." In directed graphs may consider *in-degree* the number of edges sent TO a node and *out-degree* the number of edges sent FROM a node or the total degree which is their sum.

We can find degree in igraph and we can also turn to our adjacency matrix to see how this works.

degree(g) creates a nodal undirected (or total) degree vector.

```{r results='hide', echo=TRUE, message=FALSE}
valente <- read.graph("data/valente.net", format="pajek")
gender <- as.matrix(read.table("data/valente.clu", skip=1))	
V(valente)$gender <- as.vector(gender)					
is.directed(valente)
v_deg <- degree(valente)								
```

In-degree:
degree(g, mode="in")

Out-degree:
degree(g, mode="out")

```{r message=FALSE, echo=TRUE}
v_indeg <- degree(valente, mode="in")					
v_outdeg <- degree(valente, mode="out")				
#Sum to check
v_deg==v_indeg + v_outdeg								
```

## Degree Distribution

Let's see how degree is distributed in the Valente data by plotting the distribution using ggplot2. Ggplot2 is helpful for a bunch of graphic needs notes that there is a recipe/grammar to constucting graphs. 

You begin with data and add the "aesthetic mapping" or what you want to plot and then add layers and other specifications.

A generic histogram might look like this:

ggplot(data=df, aes=(variable)) + geom_histogram()


```{r message=FALSE, echo=TRUE}
library(ggplot2)

valente.degree <- data.frame(v_deg, v_indeg, v_outdeg)

ggplot(data=valente.degree, aes(v_deg)) + 
  
  geom_histogram(breaks=seq(2,14, by=1),
                col="black", 
                fill="black", 
                alpha = .2)+
 labs(title="Total Degree Distribution") +
  labs(x="Degree", y="Count")
```

We can look at the distribution of in-degree.

```{r }
ggplot(data=valente.degree, aes(v_indeg))  +
  geom_histogram(breaks=seq(-1,14, by=1), col="black", fill="black", alpha = .2)+
 labs(title="In-Degree Distribution") +
  labs(x="In-Degree", y="Count")
```

And the distribution of out-degree.

```{r message=FALSE, echo=FALSE}

ggplot(data=valente.degree, aes(v_outdeg)) + geom_histogram(breaks=seq(-1,14, by=1),
                                                          col="black", 
                                                          fill="black", 
                                                          alpha = .2)+
 labs(title="Out-Degree Distribution") +
  labs(x="Out-Degree", y="Count")
```

We can compare in-degree and out-degree.

```{r message=FALSE, echo=TRUE, fig.show='hide'}
ggplot(data=valente.degree,aes(v_indeg,v_outdeg))+
  geom_point(aes(colour=factor(V(valente)$gender+1)))+
  geom_smooth(method='lm', level=.95, formula=y~x)+
  theme(panel.background = element_blank())
```

## Network Composition

We can construct a variety of measures of network composition. For example we might just want the percentage of alters with a particular characteristics.

This is simply: (the # of alters with x)/degree. The mean of this simple composition is often reported for the entire network/data set. You can use matrices to construct these measures. 

For gender and the Valente data it would look like this: ADJMAT[32,32] X GENDER[32,2] = RECEIVE&SENT[32,2] 


```{r , echo=TRUE}

adjmat <- as_adjacency_matrix(valente)


girl <- ifelse(gender==1,1,0)
boy <- ifelse(gender==0, 1, 0)

gendermat <- cbind(girl, boy)

multiplication <- adjmat %*% gendermat

percent.cat <- data.frame(toboy=multiplication[,2], togirl=multiplication[,1], gender=gender, outgree=v_outdeg)

percent.cat$prop.boy <- percent.cat$toboy/percent.cat$outgree

percent.cat$prop.girl <- percent.cat$togirl/percent.cat$outgree

mean(percent.cat$prop.boy, na.rm=TRUE)

mean(percent.cat$prop.girl, na.rm=TRUE)


```

Rather than calling and external function, here I'm going to write one "on the fly" that we want to use. This one will get the mean of an attribute among one's alters (you could conceivably put in another function, like range or standard deviation if your question so warrants.)

```{r}
alter_mean <- function(graph, attribute, mode="all"){				# mode can also be "in" or "out"
	mean_attr <- list(0)											# initializing a list
	for (i in 1:vcount(graph)){										# for all nodes in school
		ineighbors <- neighbors(graph, i, mode)						# find the neighbors
		attr <- get.vertex.attribute(graph, attribute, ineighbors)	# finding the neighbors' attribute
		mean_attr[[i]] <- mean(attr, na.rm=T)						# averaging over the list
	}
	return(unlist(mean_attr))										# returning it as a vector
}
```

Note: in running this code chunk, all we've done is defined the new function. Now we can use it like we have others.

```{r}
gender_mean_rcv <- alter_mean(valente, "gender", "in")

mean(gender_mean_rcv) #taking the average across all observations
mean(gender_mean_rcv, na.rm=T) #removing those who receive no ties.

gender_mean_snd <- alter_mean(valente, "gender", "out")
mean(gender_mean_snd, na.rm=T) #taking the average across all observations

mean(alter_mean(valente,"gender"))
```

## Network Composition: Mixing Matrix

We may want to know how many in-group or out-group ties there are. In sociology, this offers a good indication of social closure. We may also think about it as one indication of network segregation.

```{r}
plot.igraph(valente, vertex.color=V(valente)$gender*2)
mixingm(valente, 'gender')	# the density option would provide proportional representation
```

Now let's compare to a random graph.

```{r}
rg <- erdos.renyi.game(length(V(valente)), length(E(valente)), type="gnm", directed=T)	# a random graph with the same number of nodes & edges as the Valente graph above

V(rg)$gender <- sample(c(0,1), length(V(rg)), replace=T, prob=c(22/37, 15/37))	#randomly assigning gender, matching the probabilities from the Valente graph

plot(rg, vertex.label= NA, edge.arrow.size=0.02,vertex.size = 5, vertex.color=V(rg)$gender*2, xlab = "Erdos-Renyi Random Graph")


mixingm(rg, 'gender')	# If that’s repeatedly highly assortative, we’re in trouble (occasionally it will be)!
```

Note that your results will differ from mine. These are random draws.

We can compare the Valente graph and the random graph based on assortativity - measure of homophily developed by Mark Newman that indexes the number of edges within group over the variance as a whole. It is basically Pearson's correlation coefficient and similarly varies from perfectly disassortative (-1) to perfectly assortative (1).

See:

Newman, M. E. (2003). Mixing patterns in networks. Physical review E, 67(2), 026126.

```{r, echo=TRUE}

assortativity(valente, V(valente)$gender)

assortativity(rg, V(rg)$gender)

```

## Network Composition: IQV

The Index of Qualitative Variation measures variation among nominal variables. Here, it captures the diversity across ego's alters on a given nominal characteristic. We can think of variation as a product of normalized sum of proportions. So higher scores will translate to greater variation from 0 (or no variation) to 1 (maximal variation). 

Following Marsden (1987) and many others we use Argesti and Argesti (1978:208):

*(1-Sum(Category Proportions)^2)(# of Categories/1-# of Categories)*

See:

Marsden, P. V. (1987). Core discussion networks of Americans. American sociological review, 122-131.

Agresti, A., & Agresti, B. F. (1978). Statistical analysis of qualitative variation. Sociological methodology, 9, 204-237.

```{r, echo=TRUE}
source("scripts/jimi_iqv.r")
v_iqv_gender <- as.data.frame(iqv(valente, "gender"))
quantile(v_iqv_gender, na.rm=TRUE)
mean(v_iqv_gender$egonet)

```


## Network Density

Network density provides an indication of the strong ties within the network.

It is the number of ties/number of possible ties. We can find the measure for the entire graph.

```{r, echo=TRUE}

graph.density(valente)

```

We can also compute the density for each ego network within a graph.

If you have an igraph object, you can pull out each ego network using make_ego_net.

```{r, echo=TRUE}

valente_ego_dens <- make_ego_graph(valente, 1) %>%
vapply(graph.density, numeric(1))

head(valente_ego_dens)

```

## Transitivity

Transitivity is a measure of local clustering. It consists of the proportion of closed triads given all possible triads. The igraph default is for each node but you can also calculate for the whole graph. This is an indication of how much local clustering is occuring at the graph as a whole.


```{r, echo=TRUE}

transitivity(valente)

transitive.valente <- transitivity(valente, type="local")
ego.size <- ego_size(valente)

tr.df <- data.frame(transitivity=transitive.valente, size=ego.size)

```

We can plot transitivity by ego network size.


```{r message=FALSE, echo=FALSE}
ggplot(data=tr.df,aes(transitivity,size))+
  geom_point()+
  geom_smooth(method='lm', level=.95, formula=y~x)+
  theme(panel.background = element_blank())
```

And plot transitivity as node size in a network.

```{r message=FALSE, echo=FALSE}
plot.igraph(valente, vertex.size=transitive.valente*20, edge.arrow.size=.5)
```

## Constraint

Another local measure is the extent to which an actor may be constrained by local clustering. Constraint captures the extent to which a node's local network is redundant. In other words, small networks with high redundancy will have high constraint scores. The formula itself has been called "daunting," but translates to a combination of size, density, and hierarchy.

"Network constraint measures the extent to which your time and energy are concentrated in a single group" (Burt)

See:

Burt, R. S. (2015). Reinforced structural holes. Social Networks, 43, 149-161.

Burt, R. S. (1997). A note on social capital and network content. Social networks, 19(4), 355-373.

```{r, echo=TRUE}

constraint.valente <- constraint(valente)
ego.size <- ego_size(valente)

con.df <- data.frame(constraint=constraint.valente, size=ego.size)

```

We can look at the relationship between size and constraint.

```{r message=FALSE, echo=FALSE}
ggplot(data=con.df,aes(constraint,size))+
  geom_point()+
  geom_smooth(method='lm', level=.95, formula=y~x)+
  theme(panel.background = element_blank())
```
And plot as node size.

```{r message=FALSE, echo=FALSE}
l <- layout_with_kk(valente)

plot.igraph(valente, vertex.size=con.df$constraint*10, edge.arrow.size=.01, layout=l)
```

The inverse of constraint is one way to think about brokerage. The maximal number of Burt's constraint is 1.125. So, if we subtract 1.125 from constraint we have one brokerage score.

```{r message=FALSE, echo=FALSE}
con.df$inv <- 1.125-con.df$constraint

plot.igraph(valente, vertex.size=con.df$inv*10, edge.arrow.size=.5)
```

## Small Worlds

Small worlds graphs have nodes with high transitivity -triads tend to close - and short average path length overall. The shortest path length, or geodesic, is the fewest number of edges that a thing would have to traverse to get from one node to another. You can use the shortest.paths function to get the matrix of shortest paths. A small world graph is the kind of graph that we would expect given the strength of weak ties hypothesis. Recall that transitivity of the graph can be found by transitivity(g).

We can build a random small world graph using the watts.strogatz.game and compare to the previous simulated graph and the Valente graph.

See: 
Watts, D. J., & Strogatz, S. H. (1998). Collective dynamics of ‘small-world’networks. nature, 393(6684), 440-442.

Watts, D. J. (2004). Small worlds: the dynamics of networks between order and randomness. Princeton university press.

 
```{r }

sim <- watts.strogatz.game(1, 37, mean(degree(valente)), 0.5, loops = FALSE, multiple = FALSE)

plot(sim, vertex.label= NA, edge.arrow.size=0.02,vertex.size = 5, xlab = "Small world model")


mean(shortest.paths(sim))

mean(shortest.paths(rg))

mean(shortest.paths(valente))

transitivity(sim)

transitivity(rg)

transitivity(valente)

```

