---
title: "HW6 Cheat Sheet"
format: 
  html:
    code-fold: true
    embed-resources: true
execute:
  warning: false
editor: source
---

This is a cheat sheet for the sixth homework on cohesion and community. It works through each step in R.

First, you need the following packages.

```{r, message=FALSE}
library(igraph)
library(ggplot2)
library(Hmisc)
```

Next, we bring the data into **igraph**. We have an igraph object stored as a .rds file. The data consist of an advice seeking network within a consulting firm: (“Please indicate how often you have turned to this person for information or advice on work-related topics in the past three months”). 0: I Do Not Know This Person; 1: Never; 2: Seldom; 3: Sometimes; 4: Often; and 5:Very Often.) Not knowing and never seeking advice edges have been dropped. The **igraph** objects include two attributes - gender (1: male; 2: female), region (1: Europe; 2: USA). You can learn more about the data [here](https://toreopsahl.com/datasets/#Cross_Parker).

```{r, message=FALSE}

orgnet <- readRDS("data/orgnet.rds")

summary(orgnet)

plot(orgnet, vertex.color=V(orgnet)$region, edge.arrow.size=.1)

```

```{r, message=FALSE}

bicomponent.org <- biconnected.components(orgnet)

b1 <- bicomponent.org$components[[1]]

bi.g <- induced.subgraph(graph=orgnet,vids=b1)

plot.igraph(bi.g, vertex.color=V(orgnet)$region)

```

Let's graph the two networks with node color varying by adoption or not.

```{r, message=FALSE}

cliques <- largest_cliques(orgnet)

c1 <- cliques[[1]]

clique2 <- induced.subgraph(graph=orgnet,vids=c1)

plot(clique2, vertex.color=V(orgnet)$region, edge.arrow.size=.1)


plot(clique2, vertex.color=V(orgnet)$gender, edge.arrow.size=.1)


```

Here 

```{r, message=FALSE}

kcore <- coreness(orgnet)    

table(kcore)

```

Here

```{r, message=FALSE}


```


```{r, message=FALSE}
```


```{r, message=FALSE}


```


We can also construct some formal tests.

```{r, message=FALSE}


```

We could even construct a logistic regression predicting likelihood of adopting the new practice.

```{r, message=FALSE}




```

