---
title: 'Week 7: Affiliation Networks'
author: "Ryan Light with significant collaboration with jimi adams"
date: "May 12, 2021"
output:
  html_document:
    df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Affiliation Networks
How do social actors or entities “hang” together?

  -Direct ties: Social relations

  -Indirect ties: Membership
  
    - What does it mean to be affiliated through a literal group, activity, 
    or event?
 
 “A man may regarded as the point of intersection of an indefinite number of circles representing social groups, having as many arcs passing through him as there are groups.” (Cooley)

What does it mean to be connected through a generic group or an abstraction?

In this workshop, we will revisit ideas that have already been introduced during the visualization workshop and introduce the new academictwitteR package.

## Duality: Breiger (1974)

The classic statement on this process is Breiger's (1974) concept of duality.

Metaphor: People intersect through their associations, which defines (in part) their individuality.

Duality: Relations among groups implies relations among individuals or vice versa.

We call these persons-thorugh-groups networks two-mode networks, affiliation networks, or bipartitie networks (among others).

The consequence is that we expand the amount of data that can be "networked" as we move beyond standard (and expensive) conceptions as networks as self-reported (surveys) or observed (ethnographies) of social relations.

With most social media, we can observe social relations of the standard type in the form of of followers and followed networks, but also two-mode netwokrs in the form fo retweet networks, hashtag networks, co-word networks, etc.

Here we will look at a hashtag networks on Twitter using the academictwitteR package.

## Let's set up acaedmictwitteR.

First we have to set up an academic developer account with Twitter that will allow for the authorization required to work with Twitter's API (application programming interface). An API allows two applications to talk to each other. In this case and in many others, Twitter has used an API to allow for the distribution of some data. Very recently Twitter expanded access to their data to academics and so they make you fill out some information to apply for this access, including confirming your university affiliation.

https://cran.r-project.org/web/packages/academictwitteR/vignettes/academictwitteR-auth.html

This process can take a day or two for confirmation.

In the end you will get a "bearer token" that will allow the academictwitteR package to query Twitter.

This token is a personal thing, so you have to grab your own. For the workshop we will start with the bits after data are collected.

```{r, echo=T, eval=FALSE}
#load the package

library(academictwitteR)

#Use your token here

bearer_token <- "YOUR TOKEN" # Insert bearer token


```

Now we can build a query to ask twitter for data on a particular term or user. Here, we show are interested in hashtags and users.

We will set up a query that skips over retweets:

```{r, echo=T, eval=FALSE}

#List hashtags

htagquery <- c("#socAF", "#soctwitter")

#List query elements. Here getting rids of retweets

no_retweets <- build_query(htagquery, is_retweet=FALSE)

```

We will use the get_all_tweets function with this query and store the downloaded data as a json file as recommended due to potential download disruptions. The data_path argument tells us where this is being stored.

We then use the bind_tweet_jsons function to push those files into a data.frame.

```{r, echo=T, eval=FALSE}

#This gets the tweets

soc_json <-
  get_all_tweets(
    no_retweets,
    "2021-05-01T00:00:00Z",
    "2021-05-10T00:00:00Z",
    bearer_token, 
    data_path = "data/",
    bind_tweets = FALSE
  )

```

We then use the bind_tweet_jsons function to push those files into a data.frame.

```{r, echo=T}

library(academictwitteR)

#Here we push the json files together

tweets <- bind_tweet_jsons(data_path = "data/")

```

Next we pull out a list of authors because we would like to construct a two-mode network of users and hashtags.

The tweets data frame will have an author_id. We can save that as a vector and use that for the get_user_profile function that asks the Twitter API for data on specific users.

```{r, echo=T, eval=FALSE}

#Here we grab the list of users

users <- tweets$author_id

users <- get_user_profile(users, bearer_token)

save(users, tweets, file="twitter_data.rda")

```


## We have twitter data now how do we make a two-mode hashtag network.

In the tweets file we can see that their are multiple things stored as entities. One of the entities is a  list of dataframes of hashtags for each tweet. You have to inspect the data frame to see which of the entities captured the hashtags.

So we are going to pull that out and unlist using the unnest function from dplyr.

```{r, echo=T}

#Load Twitter user and tweets data using RStudio.

load("~/Documents/GitHub/sna_course/week_8_affiliation/twitter_data.rda")


library(dplyr)
library(tidyr)

#This jumbles on each download so you have to inspect data frrame

ents <- data.frame(tweets$id, tweets$entities[3])


hashs <- ents %>% unnest(hashtags) %>% 
  group_by(tweets.id)


```


Now we will use left_join from dplyr to connect the authors to hashtags by the id for each tweet that they have in common.

This will make an author by hashtag edge list.

```{r, echo=T}

twt <- data.frame(tweets.id = tweets$id, auid = tweets$author_id)

hash_author <- left_join(hashs, twt)

hash_edge <- data.frame(auid=hash_author$auid, tag=hash_author$tag)

hash_edge$tag <- trimws(tolower(hash_edge$tag))

head(hash_edge)

```

Now we are back to where we have been with igraph. We will use graph.data.frame to bring in the edege list.

This first graph won't be identified as a two-mode network. We have to set up the type (V(g)$type) so igrpah can tell that it is a two-mode network. 

The bipartite_mapping function identifies the two distinct types. 

```{r, echo=T}

library(igraph)

tw.g <- graph.data.frame(hash_edge)


V(tw.g)$type <- bipartite.mapping(tw.g)$type

```

We can set some parameters to mae the graph pretty and to better identify the modes and then plot the graph.

```{r, echo=T}
V(tw.g)$color <- ifelse(V(tw.g)$type, "lightblue", "salmon")
V(tw.g)$shape <- ifelse(V(tw.g)$type, "circle", "square")
E(tw.g)$color <- "lightgray"

plot(tw.g, vertex.size=4, edge.arrow.size=.01, layout=layout.fruchterman.reingold)

```


It is pretty messy so let's look at it without labels.


```{r, echo=T}

plot(tw.g, vertex.size=4, edge.arrow.size=.01, vertex.label="", layout=layout.fruchterman.reingold)

```

We know by defintion that socaf and soctwitter are likely to be the center of the graph.

So it makes sense to use delete.vertices to remove them. We can also label a set of nodes and redraw.


```{r, echo=T}

V(tw.g)$name <- trimws(tolower(V(tw.g)$name))

tw.g2 <- delete.vertices(tw.g, V(tw.g)$name=="socaf")

tw.g2 <- delete.vertices(tw.g2, V(tw.g2)$name=="soctwitter")


lb.df <- data.frame(type=V(tw.g2)$type, name=V(tw.g2)$name)

lb.df$nms <- ifelse(lb.df$type=="TRUE", lb.df$name, " ")

plot.igraph(tw.g2, layout=layout.fruchterman.reingold, vertex.label.color="black", vertex.label=lb.df$nms, vertex.size=4, edge.arrow.size=.01)

```

Still messy...let's just look at that main component as things fractured with the hashtag delete.

```{r, echo=T}

comp.g <- induced_subgraph(tw.g2, components(tw.g2)$membership==1)

isos <- which(degree(comp.g)==0)

comp.g <- delete.vertices(comp.g, isos)

lb.df <- data.frame(type=V(comp.g)$type, name=V(comp.g)$name)

lb.df$nms <- ifelse(lb.df$type=="TRUE", lb.df$name, " ")

plot.igraph(comp.g, layout=layout.fruchterman.reingold, vertex.label.cex=.5, vertex.label.color="black", vertex.label=lb.df$nms, vertex.size=4, edge.arrow.size=.01)
```

As you can probably imagine two-mode networks require some alternative measurement approaches for thinking about centrality, community and so forth. This is still a relatively new area, but there is a package called tnet that is useful for this kind of analysis.

## Bipartite Projections

However, most social scientists will still project the two-mode network to a one-mode network (either persons connected to persons via common groups or groups connected to groups via common persons).

The bipartite.projection function does just that. The result is a list of graphs. One for the first mode and one for the second.


First, let's look at the hashtags or the "groups" here.



```{r, echo=T}

bip.g <- bipartite.projection(comp.g)

hash.g <- bip.g$proj2

V(hash.g)$deg <- degree(hash.g, normalized=TRUE)

hashlab <- data.frame(name=V(hash.g)$name, degree=V(hash.g)$deg)

hashlab$lab <- ifelse(hashlab$degree>.13, hashlab$name, " ")

plot.igraph(hash.g, layout=layout.fruchterman.reingold, vertex.label.cex=.5, vertex.label.color="black", vertex.label=hashlab$lab, vertex.size=4, edge.arrow.size=.01)

```

Still pretty messy. Let's see if we can cut down a bit more.

```{r, echo=T}

nhash.g <- delete.edges(hash.g, which(E(hash.g)$weight<2))

isos <- which(degree(nhash.g)==0)

nhash2.g <- delete.vertices(nhash.g, isos)

plot.igraph(nhash2.g, layout=layout.fruchterman.reingold, vertex.label.cex=.5, vertex.label.color="black", vertex.size=4, edge.arrow.size=.01)


```

And of course we can do the same thing with the twitter authors projection.

```{r, echo=T}

peep.g <- bip.g$proj1

plot.igraph(peep.g, layout=layout.fruchterman.reingold, vertex.label.cex=.5, vertex.label="", vertex.label.color="black", vertex.size=4, edge.arrow.size=.01)

```





Again these one-mode projections are subject to the same analyses as other "traditional" person-to-person networks that we've discussed.

```{r, echo=F}


#library(knitr)
#purl("w7_affilliation_workshop.Rmd")

```